Two applications of logic programming to Coq

### 2021-02-22 papers/types2020/revisions/pbt.tex
cleanup of the code directoty under types2020

Of course, Coq already features \textsf{QuickChick}~\cite{QChick}
(\url{https://softwarefoundations.cis.upenn.edu/qc-current}), which is
a sophisticated and well-supported PBT tool, based on a different
perspective: being a clone of Haskell's QuickCheck, it emphasizes
testing executable (read \emph{decidable}) specifications with random
generators. While current research~\cite{LampropoulosPP18} aims to
increase automation, it is fair to say that testing with
\textsf{QuickChick}, in particular of relational specifications, is
still very labor intensive. We do not intend to compete with
\textsf{QuickChick} at this stage, but we shall see that we can test immediately
\lsti{Inductive} definitions that corresponds to pure Horn
programs, without having to provide a decidability
proof for those definitions. Furthermore, we are not committed to a
fixed random generation strategy, which in general requires additional
work in the configuration of generators and shrinkers.

@@

To exemplify, let us add to the previous specification of list
insertion a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x :: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is
whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall (x : nat) xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to height
at most $5$  from the library \lsti{Inductive} definitions of
\lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
since by mode information we know that it will be computed.
%
Since we did slip in an error, our tactic reports a counter-example,
namely \verb|Proof Term: [0, [0; 1; 0]]|, which unpacks to \lsti{x = 0}
and \lsti{xs = [0; 1; 0]}. As the latter is definitely not an
ordered list, this points to a quite evident bug in the definition of
\lsti{ordered}. We leave it to the reader and we abort the proof.

@@

Another way to asses the fault detection capability of a PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are voluntarily inserted, with the view that they should caught (``killed'') by a ``good enough'' testing suite. Following on  an exercise in the aforementioned chapter of SF, we modify the typing relation by adding the following (nonsensical)  clause:

\begin{lstlisting}
Module M1.
Inductive has_type : tm -> typ -> Prop :=
$\dots$
  | T_SuccBool : forall t, has_type t TBool -> has_type (tsucc t) TBool.
end M1.
\end{lstlisting}
Some of the required properties for our SUT now fails: not
only type uniqueness, but also progress with counterexample \lsti{e = tsucc ttrue}:
\begin{lstlisting}
Definition progress (e :tm) (Has_type : tm -> typ -> Prop) (Step : tm -> tm -> Prop):= 
     forall t, Has_type e t -> notstuck e Step.
Conjecture progress_m1:  forall e, progress e M1.has_type step.
 unfold progress.
 intros e t Ht.    
 elpi dep_pbt height 2 (Ht) (e) .
 Abort.
\end{lstlisting}

>>>>

Of course, Coq already features \textsf{QuickChick}~\cite{QChick}
(\url{https://softwarefoundations.cis.upenn.edu/qc-current}), which is
a sophisticated and well-supported PBT tool, based on a different
perspective: being a clone of Haskell's QuickCheck, it emphasizes
testing executable (read \emph{decidable}) specifications with random
generators. While current research~\cite{LampropoulosPP18} aims to
increase automation, it is fair to say that testing with
\textsf{QuickChick}, in particular relational specifications, is
still very labor intensive. We do not intend to compete with
\textsf{QuickChick} at this stage, but we shall see that we can test immediately
\lsti{Inductive} definitions that corresponds to pure Horn
programs, without having to provide a decidability
proof for those definitions. Furthermore, we are not committed to a
fixed random generation strategy, which in general requires additional
work in the configuration of generators and shrinkers.

@@

To exemplify, let us add to the previous specification of list
insertion a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x :: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is
whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall (x : nat) xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to height
at most $5$  from the library \lsti{Inductive} definitions of
\lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
since by (informal) mode information we know that it will be computed.
%
Since we did slip in an error, our tactic reports a counter-example,
namely \verb|Proof Term: [0, [0; 1; 0]]|, which unpacks to \lsti{x = 0}
and \lsti{xs = [0; 1; 0]}. As the latter is definitely not an
ordered list, this points to a quite evident bug in the definition of
\lsti{ordered}. We leave the fix to the reader. % and we abort the proof.

@@

Another way to asses the fault detection capability of a PBT setup is via \emph{mutation analysis}~\cite{CavadaCM20}, whereby localized bugs are purposely inserted, with the view that they should caught (``killed'') by a ``good enough'' testing suite. Following on  an exercise in the aforementioned chapter of SF, we modify the typing relation by adding the following (nonsensical)  clause:

\begin{lstlisting}
Module M1.
Inductive has_type : tm -> typ -> Prop :=
$\dots$
  | T_SuccBool : forall t, has_type t TBool -> has_type (tsucc t) TBool.
end M1.
\end{lstlisting}
Some of the desired properties for our SUT now fails: not
only type uniqueness, but also progress with counterexample \lsti{e = tsucc ttrue}:
\begin{lstlisting}
Definition progress (e : tm) (Has_type : tm -> typ -> Prop) (Step : tm -> tm -> Prop):= 
     forall t, Has_type e t -> notstuck e Step.
Conjecture progress_m1:  forall e, progress e M1.has_type step.
 unfold progress.
 intros e t Ht.    
 elpi dep_pbt height 2 (Ht) (e).
 Abort.
\end{lstlisting}

### 2021-02-22 papers/types2020/revisions/sec3.tex
cleanup of the code directoty under types2020

The proof system is parameterized by Coq's global environment $E$,
which here we take as a set of inductive definitions; following Coq's
reference manual, we denote such a definition with
$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
the names and types of the (possibly mutually)inductive type and
$\Gamma_C$ the names and types of its constructors; finally $p$
denotes the number of parameters and playes no role. We delegate to
Coq's type checking the enforcement of the well-sortedness of
inductive types. The \emph{local context} is empty, since we are only
dealing with types that correspond to Horn clauses, and atomic types
are inductively defined. In fact, we do not have a $\forall$ rule on
the right, although the proof theory would gladly allow it. This means
at the term level that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.  The \emph{decide} rule, % is the equivalent in our
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
 selects a clause on which to back-chain on: we lookup the
constructors of an \lsti{Inductive} definition from the global
environment, one that matches the head symbol of the atom we aim to backchain on, and then
call the latter judgment that will find a correct instantiation, if any. The rules for backchaining are standard,
corresponding to the \emph{init} and the (conflation of the) left introduction rules for
$\forall$ and $\supset$.

@@

\subsubsection{Coq-specific code}
Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is part of the Coq-Elpi API signature of constants we use:
\begin{lstlisting}[language=lprolog]
kind term   type.                               % reification of Coq terms
kind gref   type.                               % reif. of references to global terms
type global gref -> term.                            % coercion to term
type indt   inductive -> gref.                       % reification of inductive types
type indc   constructor -> gref.                     % reification of their constructors 
type app    list term -> term.                       % reification of nary application
type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}
%
Note that \lsti{prod} encodes dependent products by taking a name for pretty
printing, a term and a \lP abstraction from terms to terms: i.e., $(x :
B) D$ is encoded by \verb|prod ``x'' B (x\ D x)|; when, in the
implementation of the product-left rule, we apply \lsti{D} to the variable
\lsti{Tm}, we get a new formula that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in the  LJT and PTSC's tradition, the more since proof search in
our application is cut-free.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The decide rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.
The \lsti{decideE} predicate,  has, among others, the role of selecting which
constructor to focus on from the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase.

@@

Thanks to the Coq-Elpi interface, in particular to the ``main''
procedure \lsti{solve}, we can package the \lP's code for the checker
as a tactic that can be called as any other tactic in a Coq script.

@@

The glue code between Coq-Elpi and the implementation of our calculus
is straightforward: the goal consists of a quadruple of a (here
inactive) context, an \emph{evar}, a type (goal) and a list of extra
information, also inactive. In addition, we supply the certificate: it
is simply an integer and a string to identify the ``resource'' FPC that we
will use in this case. We just need to call \lsti{check} on the goal,
together with the certificate, in order to obtain a reconstructed
proof term. We do not call the reconstruction directly on the evar
because Coq-Elpi ensures that evars manipulated by \lP are well-typed
at all times; since we cannot guarantee that, as we work with
partially reconstructed term, we get around this by an explicit
unification.

@@

The \lsti{dprolog} tactic can be seen as a programmable (and possibly
faster) version of Coq's \lsti{eauto}, although it currently lacks the
latter's capability of solving goals by
\lsti{reflexivity}; % as well as the modularity of using
% \lsti{Hints} as introduction rules
 \lsti{dprolog} is not restricted to depth-first search, since
it follows the dictate of the given FPC, for example we could easily add
iterative-deepening search. Furthermore, the tactic can still use
\lsti{Hints} via Elpi's \lsti{Db} feature, and FPCs can provide a
\emph{trace} that may be more customizable than the one offered by the
\lsti{(e)auto}'s hard-wired \lsti{Debug} facility.

>>>>

The proof system is parameterized by Coq's global environment $E$,
which here we take as a set of inductive definitions; following Coq's
reference manual, we denote such a definition with
$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
the names and types of the (possibly mutually)inductive type and
$\Gamma_C$ the names and types of its constructors; finally $p$
denotes the number of parameters and plays no role. We delegate to
Coq's type checking the enforcement of the well-sortedness of
inductive types. The \emph{local context} is empty, since we are only
dealing with types that correspond to Horn clauses, and atomic types
are inductively defined. In fact, we do not have a $\forall$ rule on
the right, although the proof theory would gladly allow it. This means
at the term level that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.  The \emph{decide} rule, % is the equivalent in our
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
 selects a clause on which to back-chain on: we lookup the
constructors of an \lsti{Inductive} definition from the global
environment, one that matches the head symbol of the atom we aim to backchain on, and then
call the latter judgment that will find a correct instantiation, if any. The rules for backchaining are standard,
corresponding to the \emph{init} and the (conflation of the) left introduction rules for
$\forall$ and $\supset$.

@@

\subsubsection{Coq-specific code}
Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of native \lP constructs such as lists and binders.
The following is part of the Coq-Elpi API signature of constants that we use:
\begin{lstlisting}[language=lprolog]
kind term   type.                                 % reification of Coq terms
kind gref   type.                                 % reif. of references to global terms
type global gref -> term.                            % coercion to term
type indt   inductive -> gref.                       % reification of inductive types
type indc   constructor -> gref.                     % reification of their constructors 
type app    list term -> term.                       % reification of nary application
type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}
%
Note that \lsti{prod} encodes dependent products by taking a name for pretty
printing, a term and a \lP abstraction from terms to terms: i.e., $(x :
B) D$ is encoded by \verb|prod ``x'' B (x\ D x)|; when, in the
implementation of the product-left rule, we apply \lsti{D} to the variable
\lsti{Tm}, we get a new term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in the  LJT and PTSC's tradition; this is consistent with proof search in
our application being cut-free, whereas explicit substitutions are linked to cuts.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The decide rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} to access the global
environment of inductive definitions and query for information about them.
The \lsti{decideE} predicate,  has, among others, the role of selecting which
constructor to focus on from the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase.

@@

Thanks to the Coq-Elpi interface, in particular to the ``main''
procedure \lsti{solve}, we can package the \lP code for the checker
as a tactic that can be called as any other tactic in a Coq script.

@@

The glue code between Coq-Elpi and the implementation of our calculus
is straightforward: the goal consists of a quadruple of a (here
inactive) context, an \emph{evar}, a type (goal) and a list of extra
information, also inactive. In addition, we supply the certificate: it
is simply an integer and a string to identify the ``resource'' FPC that we
will use in this case. We just need to call \lsti{check} on the goal  \lsti{G},
together with the certificate, in order to obtain a reconstructed
proof term. We do not call the reconstruction directly on the evar
because Coq-Elpi ensures that evars manipulated by \lP are well-typed
at all times; since we cannot guarantee that, as we work with
partially reconstructed term, we get around this by an explicit
unification.

@@

The \lsti{dprolog} tactic can be seen as a programmable (and possibly
faster) version of Coq's \lsti{eauto}, although it currently lacks the
latter's capability of solving goals by
\lsti{reflexivity}; % as well as the modularity of using
% \lsti{Hints} as introduction rules
 \lsti{dprolog} is not restricted to depth-first search, since
it follows the dictate of the given FPC; for example we could easily add
iterative-deepening search. Furthermore, the tactic can still use
\lsti{Hints} via Elpi's \lsti{Db} feature, and FPCs can provide a
\emph{trace} that may be more customizable than the one offered by 
\lsti{(e)auto}'s hard-wired \lsti{Debug} facility.
